Secure Composition of Robust and Optimising Compilers

### 2023-07-09 popl/main.tex
feedback on 4

This section presents the key meta-theoretic results of this paper and builds on prior work~\cite{kruse2022csc}\MK{mention this at all? may reveal identity even though its double-blind}.
A main result is that secure compilers in the robust preservation framework~\cite{abate2019jour} compose.
This is not intuitive in the sense that there are a number of examples in the security domain where composition does not work~\cite{canetti2006univcomp,mccullough2012compo,fabian2022automatic}.
For starters, the sequential composition of compilers $\cc{\src{L}}{\trg{L}}$ and $\cc{\trg{L}}{\obj{L}}$ is as expected, i.e., given an $\src{L}$ program $\src{p}$, obtain its compiled $\obj{L}$ program by plugging $\src{p}$ into $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$.
Now, given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\trg{L}}{\obj{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their sequential composition $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.

@@

Notably, real-world compilation chains also perform a series of passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in the wild~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion}, it is infeasible to list them all.
Consider the compilation passes of LLVM\cite{lattner2004llvm} whose source-code optimisations are industry--competitive.
In LLVM, the pass manager can be used to schedule passes in a way where the optimisations yield the best results.
Take \gls{cf} and \gls{dce}.
Unfortunately, the order in which \gls{cf} and \gls{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration.
Compiler engineers typically try to find an order of optimisations that yields well--optimised programs for either code size~\cite{cooper1999geneticphases} or performance~\cite{kulkarni2006exhaustivephase}.
The following corollary justifies that any such order of compilation passes is valid with respect to security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.

@@

Besides the sequential composition, there are two other compositions that are more at the language- than the compiler--level.
Nevertheless, consider $\trg{JVM BC}$, which is a popular target for programming language designers due to its high performance and relevance in industry.
As an example, $\src{Java}$ and $\obj{Kotlin}$ compile to $\trg{JVM BC}$.
Technically speaking, they both lower to class files and $\obj{Kotlin}$ objects are considered to be the same as $\src{Java}$ objects at that point.
A compiler that accepts both $\src{Java}$ and $\obj{Kotlin}$ code translating to the same target language, or rather, intermediate representation, performs a kind of {\em upper} composition.
Given a secure compiler from $\src{Java}$ to $\trg{JVM BC}$ and one from $\obj{Kotlin}$ to $\trg{JVM BC}$, it is easy to envision that one can engineer a combined compiler that accepts both $\src{Java}$ and $\obj{Kotlin}$.
A practical example of this, which is, however, insecure, is {\tt Android Studio}~\cite{androidstudio}.
The proof of \Cref{thm:rtp} can be easily adapted to also works for this {\em upper} composition.
The corresponding theorem is not listed, since there is no general definition of this kind of composition.

@@

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls{cf} and \gls{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls{cf} or \gls{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the righthandside of the figure.}\label{fig:cfdceex}
\end{figure}

%\section{Case Study: Memory Safe and Cryptographic Constant Time Code}\label{sec:casestud}
\section{Case Study: Preliminary Definitions\pages{2}}\label{sec:casestud:defs}

This section defines programming languages that the secure compilers defined in the next section will make use of.
To this end, this section defines the languages $\Ltms$, $\Ltrg$, $\Lms$, and $\Lscct$.
In this paper, $\Ltms$ is the only statically typed language and exhibits the property that all well-typed programs are \gls{tms}.
However, $\Ltms$ programs are not \gls{sms}.
That is, there are well--typed $\Ltms$ programs that perform an out--of--bounds access.
In contrast, $\Ltrg$ is untyped and does not provide any guarantees with regards to \gls{ms}.
$\Lms$ is exactly the same language as $\Ltrg$, but this paper still distinguishes the two for sake of readability.
All three languages --- so $\Ltms$, $\Ltrg$, and $\Lms$ --- assume \gls{cct} to hold.
Even though it is possible to write programs that, e.g., return early in loops, which is trivially not \gls{cct}, they can be transformed to adhere to \gls{cct}~\cite{cauligi2019fact}.
The language $\Lscct$ allows violating \gls{cct} by emitting events on, e.g., branching and division, that contain secrets.
As a countermeasure, the language provides a way to read and write to a {\em model--specific register} that enables a ``\gls{cct}-mode''.

\subsection{Shared Language Definitions}\label{subsec:cs:defs}

\begin{gather*}
  \begin{aligned}
  \mi{(Expressions)}~\expr&\bnfdef\ x \mid \valueexpr \mid \lbinop{\expr[_{1}]}{\expr[_{2}]} \mid \lget{x}{\expr} \mid \lnew{\expr[_{1}]}{\expr[_{2}]} \mid \ldelete{x} \mid \lset{x}{\expr[_{1}]}{\expr[_{2}]} \\
    & \mid \llet{x}{\expr[_{1}]}{\expr[_{2}]}  \mid \lreturn{\expr} \mid \lcall{g}{\expr} \mid \lifz{\expr}{\expr[_{1}]}{\expr[_{2}]} \mid \labort \\
  \mi{(Symbols)}~&\function\bnfdef\ \lfunction{g}{x}{e}\hspace{0.5cm}
  \mi{(Libraries)}~\library\bnfdef\ \hole{\cdot} \mid \function,\library \\
  \mi{(Programs)}~&\prog{\library_{\ctx}}{\library_{\comp}}\hspace{0.5cm}
  \mi{(Substitutions)}~\substvar\bnfdef \hole{\cdot} \mid \subst{\valueexpr}{x},\gamma \\
  \end{aligned}
\end{gather*}

Above is a fraction of the shared syntax of all the programming languages of this paper.
The syntax and semantics are mostly standard.
However, when allocating pointers with $\lnew{\expr[_{1}]}{\expr[_{2}]}$, the expression $\expr[_{1}]$ determines the initial value of the objects and $\expr[_{2}]$ the overall size of an array.
The data associated to pointers resides in the dynamic state $\Omega$ that contains sandboxed heaps, one for the context and the other for the component.
Because of the sandboxing, locations pointing into heaps must be associated with a control--tag to know the heap they point into.
Other metadata in $\Omega$ entails a list of callable objects with information on wether they belong to the context or the component.
That way, when calling a callable object, the semantics can take care of emitting appropriate events that signal a context switch.
For the languages considered in this paper, no callable objects are first--class.
So, the only values are pairs, pointers, and natural numbers.
Another non--standard part of the languages is the concept of the poison--tag.
The poison--tag marks a location as freed ($\poisoned$) without removing the information associated to the freed pointer from $\Omega$.
This measure prevents the semantics from getting stuck when an already freed pointer is freed again.
Because of this, it is possible that programs like $\llet{x}{\lnew{0}{5}}{\ldelete{x};\ldelete{x}}$ reduce and emit a trace along the lines of $\ev{Alloc\ \loc\ \sandboxtag\ 5}\cdot\ev{Dealloc\ \loc\ \sandboxtag}\cdot\ev{Dealloc\ \loc\ \sandboxtag}$.

\begin{gather*}
  \begin{aligned}
  \mi{(Pre\text{--}Events)}~\event[_{b}]\bnfdef \ev{Alloc}\ \loc\ \valueexpr \mid \ev{Dealloc}\ \loc \mid \ev{Get}\ \loc\ \valueexpr &\mid \ev{Set}\ \loc\ \valueexpr\ \valueexpr['] \mid \ev{Call}\ \comm\ g\ \valueexpr \mid \ev{Ret}\ \comm\ \valueexpr \mid \ev{Start} \mid \ev{End}\ \valueexpr \\
  \mi{(Events)}~\event&\bnfdef \emptyevent \mid \lightning \mid \tup{\event[_{b}]}{\sandboxtag} \\
  \end{aligned}
\end{gather*}

@@

\subsection{$\Lscct$: An Unsafe Language}\label{subsec:lscct}

@@

\section{Case Study: Secure Compiler Compositions in the Wild\pages{7}}\label{sec:casestud:rtp}

>>>>

This section presents the key meta-theoretic results of this paper concerning sequential compiler composition (and of optimisation passes) (\Cref{sec:sequential}) and concerning other kinds of compiler composition (\Cref{sec:other-compos}).
% and builds on prior work~\cite{kruse2022csc}\MK{mention this at all? may reveal identity even though its double-blind}. % MP: not needed at all, ever, workshop work does not need to be cited in the main work

\subsection{Secure Sequential Composition}\label{sec:sequential}

The main result is that secure compilers in the robust compilation framework~\cite{abate2019jour} compose {\em sequentially}.
This is not intuitive in the sense that there are a number of examples in the security domain where sequential composition does not work~\cite{canetti2006univcomp,mccullough2012compo,fabian2022automatic}.\MP{?canetti?}
The sequential composition of compilers $\cc{\src{L}}{\trg{L}}$ and $\cc{\trg{L}}{\obj{L}}$ is defined as follows: given an $\src{L}$ program $\src{p}$, its compiled $\obj{L}$ counterpart is obtained by plugging $\src{p}$ into $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$.
% MP: i know, i added a passive verb. here it works. i think. but i appreciate the effort!
Now, given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\trg{L}}{\obj{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their sequential composition $\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.

@@

\subsubsection{Securing Optimisations}
\MPin{note that i'm always using british spelling. optimisation VS optimization. etc. }

Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
\MPin{
	do not say the negation of something before the non-negation.
	say what things are first, and what they are not later (if at all)
}
Both examples can be seen in the wild~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion}, it is infeasible to list them all\MP{survey to cite? this is a bit informal}.
Consider the compilation passes of LLVM~\cite{lattner2004llvm} whose source-code optimisations are industry--competitive\MP{meaning?}.
In LLVM, the pass manager\MP{meaning} can be used to schedule passes in a way where the optimisations yield the best results.
Take \gls{cf} and \gls{dce}. \MPin{
	brief explanation?
}
Unfortunately, the order in which \gls{cf} and \gls{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\MPin{figure here, or closer, not far away}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration\MP{of what}.\MP{cit?}
Compiler engineers typically try to find an order of optimisations that yields well-optimised programs for either code size~\cite{cooper1999geneticphases} or performance~\cite{kulkarni2006exhaustivephase}.
\Cref{corr:swappable} justifies that any such order of compilation passes is valid with respect to security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.

@@

\subsection{Secure Upper and Lower Composition}\label{sec:other-compos}
Besides sequential composition, there are two other compositions that are more at the language- than the compiler-level\MP{namely ...}.
Nevertheless, consider $\trg{JVM BC}$, which is a popular target for programming language designers due to its high performance and relevance in industry.
As an example, $\src{Java}$ and $\obj{Kotlin}$ compile to $\trg{JVM BC}$.
Technically speaking, they both lower\MP{undefined verb. no need for synonyms here.} to class files and $\obj{Kotlin}$ objects are considered to be the same as $\src{Java}$ objects at that point.
A compiler that accepts both $\src{Java}$ and $\obj{Kotlin}$ code translating to the same target language, or rather, intermediate representation, performs a kind of {\em upper} composition.
Given a secure compiler from $\src{Java}$ to $\trg{JVM BC}$ and one from $\obj{Kotlin}$ to $\trg{JVM BC}$, it is easy to envision that one can engineer a combined compiler that accepts both $\src{Java}$ and $\obj{Kotlin}$.
A practical example of this, which is, however, insecure, is {\tt Android Studio}~\cite{androidstudio}.
The proof of \Cref{thm:rtp} can be easily adapted to also works for this {\em upper} composition.
The corresponding theorem is not listed, since there is no general definition of this kind of composition.
\MPin{
	mmm need to think about this.
	this looks like discussion section material
}
\MPin{
	we could turn this into a : "if you expected upper adn lower, they are language composition ops, not compiler ones, so that's out of scope"
}

@@

\begin{figure}
  \begin{center}
    \begin{tikzpicture}
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls{cf} and \gls{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls{cf} or \gls{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
\end{figure}

%\section{Case Study: Memory Safe and Cryptographic Constant Time Code}\label{sec:casestud}
\section{Case Study: Language Formalisations\pages{2}}\label{sec:casestud:defs}

This section defines programming languages that the secure compilers defined in the next section will make use of\MP{make use of -> use! 3 words when 1 is sufficient}.
To this end, this section defines the languages $\Ltms$, $\Ltrg$, $\Lms$, and $\Lscct$ which share many common elements (presented in \Cref{subsec:cs:defs}).
In this paper, $\Ltms$ is the only statically typed language and exhibits the property that all well-typed programs are \gls{tms} (\Cref{??}).
However, not all $\Ltms$ programs are \gls{sms}.\MP{reworded. was imprecise}
That is, there are well--typed $\Ltms$ programs that perform an out--of--bounds access.
In contrast, $\Ltrg$ is untyped and does not provide any guarantees with regards to \gls{ms} (\Cref{??}).
$\Lms$ is exactly the same language as $\Ltrg$, but this paper still distinguishes the two for sake of readability (\Cref{??}).
All three languages --- so $\Ltms$, $\Ltrg$, and $\Lms$ --- assume \gls{cct} to hold.

Even though it is possible to write programs that, e.g., return early in loops, which is trivially not \gls{cct}, they can be transformed to adhere to \gls{cct}~\cite{cauligi2019fact}.
\MPin{
	no, it's different. 
	there is no CCT because we asume it's out of the programmer's concers and since CT is arch-dependent, the compiler should preserve CT, not the programmer, via software
}

Language $\Lscct$ allows violating \gls{cct} by emitting events on, e.g., branching and division, that contain secrets (\Cref{??}).
However, $\Lscct$ also contains a countermeasure to enforce \gls{cct}, the language provides a way to read and write to a {\em model--specific register} that enables a ``\gls{cct}-mode''.
\MPin{ref to what real world stuff justifies this}

\subsection{Shared Language Definitions}\label{subsec:cs:defs}
\begin{gather*}
  \begin{aligned}
  \mi{(Expressions)}~&\expr\bnfdef\ x \mid \valueexpr \mid \lbinop{\expr[_{1}]}{\expr[_{2}]} \mid \lget{x}{\expr} \mid \lnew{\expr[_{1}]}{\expr[_{2}]} \mid \ldelete{x} \mid \lset{x}{\expr[_{1}]}{\expr[_{2}]} \\
    & \mid \llet{x}{\expr[_{1}]}{\expr[_{2}]}  \mid \lreturn{\expr} \mid \lcall{g}{\expr} \mid \lifz{\expr}{\expr[_{1}]}{\expr[_{2}]} \mid \labort
  \end{aligned}
  \\
  \begin{aligned}
  \mi{(Symbols)}~&\function\bnfdef\ \lfunction{g}{x}{e}
  &
  \mi{(Libraries)}~&\library\bnfdef\ \hole{\cdot} \mid \function,\library \\
  \mi{(Programs)}~&\prog{\library_{\ctx}}{\library_{\comp}}
  &
  \mi{(Substitutions)}~&\substvar\bnfdef \hole{\cdot} \mid \subst{\valueexpr}{x},\gamma
  \end{aligned}
\end{gather*}
\MPin{
	this latex had many issues.
	the \& were at different spots
	it used hspace to space things instead of \&
	and it did not use \& to align the following things on a line
	and it did not use multiple aligned into the gather
	(what's the point of gather if you don't have several aligned inside? use a single align* instead)
}

Above is a fraction of the shared syntax of all the programming languages of this paper.
\MP{why a fraction? what exprs are missing}
The syntax and semantics are mostly standard.\MP{and therefore omitted}
However, when allocating pointers\MP{arrays?} with $\lnew{\expr[_{1}]}{\expr[_{2}]}$, the expression $\expr[_{1}]$ determines the initial value of the objects\MP{cells?} and $\expr[_{2}]$ the overall size of an array.
The data associated to pointers resides in the dynamic state $\Omega$ that contains sandboxed heaps, one for the context and the other for the component.
Because of the sandboxing, locations pointing into heaps must be associated with a control--tag\MP{symbol} to know the heap they point into.
Other metadata in $\Omega$ entails a list of callable objects\MP{functions? objects are other things in PL} with information on wether they belong to the context or the component.
That way, when calling a callable object, the semantics can take care of emitting appropriate events that signal a context switch.
For the languages considered in this paper, no callable objects are first--class.
So, the only values are pairs, pointers, and natural numbers.
Another non--standard part of the languages is the concept of the poison--tag.
The poison--tag marks a location as freed ($\poisoned$) without removing the information associated to the freed pointer from $\Omega$.
This measure prevents the semantics from getting stuck when an already freed pointer is freed again.
Because of this, it is possible that programs like $\llet{x}{\lnew{0}{5}}{\ldelete{x};\ldelete{x}}$ reduce and emit a trace along the lines of $\ev{Alloc\ \loc\ \sandboxtag\ 5}\cdot\ev{Dealloc\ \loc\ \sandboxtag}\cdot\ev{Dealloc\ \loc\ \sandboxtag}$.

\begin{gather*}
  \begin{aligned}
  \mi{(Pre\text{-}Events)}~&\event[_{b}]\bnfdef \ev{Alloc}\ \loc\ \valueexpr \mid \ev{Dealloc}\ \loc \mid \ev{Get}\ \loc\ \valueexpr &\mid \ev{Set}\ \loc\ \valueexpr\ \valueexpr['] \mid \ev{Call}\ \comm\ g\ \valueexpr \mid \ev{Ret}\ \comm\ \valueexpr \mid \ev{Start} \mid \ev{End}\ \valueexpr \\
  \mi{(Events)}~&\event\bnfdef \emptyevent \mid \lightning \mid \tup{\event[_{b}]}{\sandboxtag} \\
  \end{aligned}
\end{gather*}

@@

\subsection{$\Lscct$: An Unsafe Language}\label{subsec:lscct}
\MPin{
	unsafe? 
	a possibly-CCT language?
}

@@

\section{Case Study: Composing Secure Compiler Passes and Optimisations \pages{7}}\label{sec:casestud:rtp}
